Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-09-22 manuscript/climate_sensitivity_draft.Rmd
finished drafting discussion!

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	## (limit 200) | 	No. of Tables | 	5  (**check if there's a figures/tables limit for regular research papers**- I don't see one)
Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
Materials and Methods |	currently ~1442	||	 
Results |	currently ~1437	 ||
Discussion |	currently ~1339	+ ~200 unwritten (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | 	~75 unwritten	||

@@

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture (Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

@@

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world (**REFS**), was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [Zuleta et al.](https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1&isAllowed=y)), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought...

@@

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height (Fig. 1a-b; see also **REFS**. **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) ([Mueller-Landau et al. 2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)). We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests.

@@

Our finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). [**DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought. ...**] Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.

Our study reveals that tree height, microhabitat, and species' hydraulic traits interactively shape tree growth responses to drought.

### new discussion outline

1. summary of main findings

2. droughts were fairly week

3. height & canopy position - biophysical factors

4. tree size - biological factors

5. we advance knowledge regarding hydraulic traits

6. we advance knowledge regarding differences across droughts

7. science is better
- by advancing our understanding of the mechanisms for individual-level responses to drought, this opens door for better predictions (elaborate from above). "This is absolutely critical to predicting forest responses to future droughts, which we're likely to see more of in the future as a result to climate change. Forecasting forest responses to these droughts is a huge and important challenge",
- furthermore, we have deepened understading of which trees will suffer , and there are implications.
- big trees

*1. paragraph summarizing main results--> primary conclusions*

We partially supported the hypothesis that crown exposure makes trees more vulnerable to drought. Co-dominant trees had the highest drought resistance. Dominant trees had lower resistance, likely because they are the most exposed. Other studies have found clear evidence of greater drought sensitivity in trees with exposed crowns (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]). At the same time, intermediate and suppressed trees had even lower resistance. This indicates that other mechanisms such as competition or rooting depth were important. (Also note that our study design was not ideal for testing the role of canopy position. Current canopy position is a conservative separator of canopy position: trees may currently be in more dominant positions than they were at the time, but backwards movement is unlikely. This would bias against finding a signficant effect for H1.2. Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)

**misc content to integrate**
From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.

our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]

No previous study has compared drought responses derived from tree-rings for so many species at one site. By comparing species at the same site, can control for other factors.

>>>>

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	## (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
Materials and Methods |	currently ~1442	||	 
Results |	currently ~1437	 ||
Discussion |	currently ~1339	+ ~200 unwritten (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | 	~75 unwritten	||

@@

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

@@

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world (**REFS**), was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [Zuleta et al.](https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1&isAllowed=y)), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.

@@

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) ([Mueller-Landau et al. 2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)). We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

@@

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**). Concerted measurement of leaf hydraulic traits of emerging importance (**REFS--KAT/NOBBY/LAWREN**) allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). [**DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought. ...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**). Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]. Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.

As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world (*REFS*), and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**). The later successional species at our site (*Fagus grandifolia*, ... ) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
